### 2014-05-12 text/Chapter04.tex
Introduction in Chapter 4

Successful programming language abstractions need to generalize a wide range of recurring
problems while capturing the key commonalities. These two aims are typically in opposition -- 
more general abstractions provide smaller value, while less general abstractions cannot be
used as often.

In the previous chapter, we outlined a number of systems that capture how computations
access the environment in which they are executed. We identified two kinds of systems --
\emph{flat} capturing whole-context properties and \emph{structural} capturing per-variable
properties. As we show in Chapter~\ref{ch:unified}, the systems can be unified using a single 
abstraction. This is useful when implementing and composing the systems, but such abstraction 
provides \emph{smaller value} -- \ie~its generality hides useful properties that we can see 
when we consider the systems separately. For this reason, this and the next chapter discusses 
\emph{flat} and \emph{structural} systems separately.

@@

Liveness is more similar to data-flow.

Implicit parameters (or resources) are a bit different, but important.

The closest work is Uustalu and Vene

@@

This section revisits the flat coeffect calculus, which provides a 
basis for the structural coeffect calculus that we introduce in Section~... We 
make a number of improvements -- we cover more standard language features and require additional laws
which allows a more useful equational theory.
The formulation is general, so as not to restrict the possible uses unnecessarily, but is
specialised enough to provide useful properties.

The flat coeffect calculus maintains a single annotation of the contextual requirements associated
with the variable context. It can be instantiated to capture implicit parameters, the system for 
dataflow and simple liveness analysis.

The expressions are those of the $\lambda$-calculus with \emph{let} binding; assuming $T$ ranges
over base types, types are extended as follows:
%
\begin{equation*}
\begin{array}{rcl}
e &::=& x \sep \lambda x.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
\tau &::=& T \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
\end{array}
\end{equation*}
%
where $T$ ranges over type constants. 
We discuss pairs and recursion in Sections~\ref{sec:flat-synprop} and \ref{sec:structural-properties}.
The type $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$
represents a function from $\tau_1$ to $\tau_2$ that requires additional context $\cclrd{r}$.
It can be viewed as a pure function that takes $\tau_1$ \emph{with} context $r$.

%Note that
%$\ctxtyp{r}{\tau}$ does not exist as syntactic value (although it has a semantic meaning
%discussed in Section~\ref{sec:flat-semantics}). The annotations $r$ are formed by 
%an algebraic structure discussed next.

@@

\subsection{Flat coeffect algebra}

@@

@@

@@

@@

\begin{figure}[t]
\begin{equation*}
\tyrule{var}
  {x : \tau \in \Gamma}
  {\coctx{\Gamma}{\aclrd{ 0 }} \vdash x : \tau }
\end{equation*}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma}{\aclrd{n'}} \vdash e : \tau }
  {\coctx{\Gamma}{\aclrd{n}} \vdash e : \tau }\quad\quad(\aclrd{n'} \leq \aclrd{n})
\end{equation*}
\begin{equation*}
\tyrule{app}
  {\coctx{\Gamma}{\aclrd{m}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{p}} \tau_2 &
   \coctx{\Gamma}{\aclrd{n}} \vdash e_2 : \tau_1 }
  {\coctx{\Gamma}{\textnormal{max}(\aclrd{m}, \aclrd{n} + \aclrd{p})} \vdash e_1~e_2 : \tau_2}
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma}{\aclrd{m}} \vdash e_1 : \tau_1 &
    \coctx{\Gamma, x:\tau_1}{\aclrd{n}} \vdash e_2 : \tau_2}
  {\coctx{\Gamma}{\aclrd{n}+\aclrd{m}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x:\tau_1}{\aclrd{n}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{n}} \tau_2 }
\end{equation*}

\caption{Flat coeffect calculus}
\label{fig:flat-types}
\end{figure}

% ------------------

\section{Type inference}

@@

% comonadic semantics
\newcommand{\iunit}{\mathsf{e}}
\newcommand{\imult}{\oplus}
\newcommand{\iftor}[1]{C^{#1}}
\newcommand{\iftorhat}[1]{\hat{C}^{#1}}

@@

\newcommand{\cobind}[2]{#1^\dagger_{#2}}
\newcommand{\cmerge}[0]{ \ident{m} }
\newcommand{\csplit}[0]{ \ident{n} }
\newcommand{\counit}[0]{ \varepsilon }
% identity function

% structural coeffect calculus - operation names

\todo{Update the introduction \& edit the whole chapter -- currently, this is just a direct copy
from the accepted ICALP paper on flat coeffects.}

\todo{Other extensions -- add more examples of indexed comonoads, consider recursion and 
fixed-point operator, possibly also consider partially defined operations, which lets us
express type systems that rule out some computations (rather than just static analyses). }

Understanding how programs affect their environment is a well studied area: \emph{effect 
systems}~\cite{effects-talpin-et-al} provide a static analysis of effects and 
\emph{monads}~\cite{monad-notions} provide a unified semantics to different notions of effect.  
Wadler and Thiemann unify the two approaches~\cite{monads-effects-marriage}, \emph{indexing}
a monad with effect information, and showing that the propagation of effects in an effect system 
matches the semantic propagation of effects in the monadic approach.

No such unified mechanism exists for tracking the context requirements. We use the term 
\emph{coeffect} for such contextual program properties. 
Notions of context have been previously captured using comonads
\cite{comonads-notions} (the dual of monads) 
and by languages derived from modal logic
\cite{logic-modal-reconstruction,logic-cmtt}, but these approaches
do not capture many useful examples which motivate our work. 
We build mainly on the former comonadic direction (\S\ref{sec:comonads}) and discuss the
modal logic approach later (\S\ref{sec:related}).

We extend a simply typed lambda calculus with a coeffect system based on comonads,
replicating the successful approach of effect systems and monads.

\paragraph{Examples of coeffects.}
We present three examples that do not fit the traditional approach of
\emph{effect systems} and have not been considered using the \emph{modal logic}
perspective, but can be captured as \emph{coeffect systems} (\S\ref{sec:motivation}) --
the tracking of implicit dynamically-scoped parameters (or resources),
analysis of variable liveness, and tracking the
number of required past values in dataflow computations.

\paragraph{Coeffect calculus.}
Informed by the examples, we identify a general algebraic structure for coeffects. 
From this, we define a general \emph{coeffect calculus} that unifies the motivating 
examples (\S\ref{sec:calculus}) and discuss its syntactic properties (\S\ref{sec:syntactic}).

\paragraph{Indexed comonads.}
Our categorical semantics (\S\ref{sec:comonads}) extends the work of Uustalu and Vene 
\cite{comonads-notions}. By adding annotations, we generalize comonads to \emph{indexed comonads}, 
which capture notions of computation not captured by ordinary comonads.

\section{Generalized coeffect calculus}
\label{sec:calculus}

The previous three examples exhibit a number of commonalities. We capture these in 
the \emph{coeffect calculus}. We do not overly restrict the calculus to make
it open for notions of context-dependent computations not discussed above.

The syntax of our calculus is that of the simply-typed lambda calculus
(where $v$ ranges over variables, \ident{T} over base types, and $r$ over coeffect annotations):
%
\begin{equation*}
e ::= v \sep \lambda v.e \sep e_1~e_2
\quad\quad\quad\quad
\tau ::= \ident{T} \sep \tau_1 \rightarrow \tau_2 \sep \ctyp{r}{\tau}
\end{equation*}

The type $\ctyp{r}{\tau}$ captures values of type $\tau$ in a context
specified by the annotation  $r$. This type appears only on the
left-hand side of a function arrow $\ctyp{r}{\tau_1} \rightarrow \tau_2$. 
In the semantics, $\ctyp{r}$ corresponds to some data type
(\eg{}~\ident{list} or \ident{Maybe}). Extensions
such as explicit \emph{let}-binding are discussed later
(\S\ref{sec:syntactic}).

The coeffect tags $r$, that were demonstrated in the previous section, can be generalized to a 
structure with three binary operators and a particular element.

\begin{definition}
A \emph{coeffect algebra} $(S, \imult, \vee, \wedge, \iunit)$ is a set $S$ with
an element $\iunit \in S$, a semi-lattice $(S, \vee)$, a monoid $(S, \imult, \iunit)$,
and a binary $\wedge$. That is, $\forall r,s,t\in S$:
\begin{equation}
\tag{monoid}
   r \imult (s \imult t) = (r \imult s) \imult t 
\quad\quad\quad
   \iunit \imult r = r = r \imult \iunit  \quad\quad
\end{equation}
\vspace{-1.5em}
\begin{equation}
\tag{semi-lattice}
   r \vee s = s \vee r 
\quad\quad
   r \vee (s \vee t) = (r \vee s) \vee t
\quad\quad
   r \vee r = r
\end{equation}
\end{definition}
%
The generalized coeffect calculus captures the three motivating examples (\S\ref{sec:motivation}),
where some operators of the coeffect algebra may coincide.

The $\imult$ operator represents \emph{sequential} composition; guided by the categorical
model (\S\ref{sec:comonads}), we require it to form a monoid with $\iunit$.
The operator $\vee$ corresponds to merging of context-requirements in \emph{pointwise composition} 
and the semi-lattice $(S, \vee)$ defines a partial order: $r \leq s$ when 
$r \vee s = s$. This ordering implies a sub-coeffecting rule.  
The coeffect $\iunit$ is often the top or bottom of the lattice.

The $\wedge$ operator corresponds to splitting requirements of a function body. 
The operator is unrestricted in the general system.
A number of additional laws holds for \emph{some} coeffects systems, \eg{}~semi-lattice structure of
$\wedge$ and a form of distributivity. Quite possibly, they should hold for all coeffect systems.  
We start with as few laws as possible so as not to limit possible uses of the calculus.  
We consider constrained variants that provide useful syntactic properties later (\S\ref{sec:syntactic}).

\vspace{-0.4em}
\paragraph{Implicit parameters} use sets of names $S = \mathcal{P}(\ident{Id})$ as tags with
union $\cup$ for all three operators. Variable access is annotated with $\iunit = \emptyset$
and $\leq$ is subset ordering.

\vspace{-0.4em}
\paragraph{Liveness} uses a two point lattice $S=\{ \ident{D}, \ident{L} \}$ where
$\ident{D} \sqsubseteq \ident{L}$. Variables are annotated with the top element $e = \ident{L}$ 
and constants with bottom $\ident{D}$. The $\vee$ operation is $\sqcup$ (join) and $\wedge$
and $\imult$ are both $\sqcap$ (meet).

\vspace{-0.4em}
\paragraph{Dataflow} tags are natural numbers $S = \mathbb{N}$ and operations
$\vee, \wedge$ and $\imult$ correspond to \textit{max}, \textit{min} and $+$, respectively. 
Variable access is annotated with $\iunit = 0$ and the order $\leq$ is the standard ordering
of natural numbers.

@@

\vspace{-1em}
\subsection{Coeffect typing rules.}
Figure~\ref{fig:coeffect-typing} shows the
rules of the coeffect calculus, given some coeffect algebra $(S, \imult, \vee, \wedge, \iunit)$.
The context required by a variable access (\emph{var}) is annotated with $\iunit$.
The sub-coeffecting rule (\emph{sub}) allows the contextual requirements of an expression 
to be generalized.

@@

In function application (\emph{app}), context requirements of both expressions and the 
function are combined as previously: the pointwise composition $\vee$ is used to combine
the coeffects of the expression representing a function $r$ and the coeffects of
the argument, sequentially composed with the coeffects of the function $s \imult t$.

@@

\begin{figure}[t]
\vspace{-1em}
\begin{equation*}
\inference[(\emph{var})\;]
  {x : \tau \in \Gamma}
  {\ctyp{\iunit}{\Gamma} \vdash x : \tau }
%
\quad\quad
%
\inference[(\emph{app})\;]
  {\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{s}{\tau_1} \rightarrow \tau_2 &
   \ctyp{t}{\Gamma} \vdash e_2 : \tau_1 }
  {\ctyp{r \vee (s \imult t)}{\Gamma} \vdash e_1~e_2 : \tau_2}
\end{equation*}
\begin{equation*}
\inference[(\emph{sub})\;]
  {\ctyp{s}{\Gamma} \vdash e : \tau }
  {\ctyp{r}{\Gamma} \vdash e : \tau }
\;(s\leq r)  
%
\quad\quad\quad\quad
%
\inference[(\emph{abs})\;]
  {\ctyp{r \wedge s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
  {\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
\end{equation*}
\caption{Type and coeffect system for the coeffect calculus}
\label{fig:coeffect-typing}
\vspace{-1em}
\end{figure}

@@

The indexed comonad structure provides a notion of composition
for computations with different contextual requirements.
%%
\begin{definition}
Given a monoid $(S, \imult, \iunit)$ with binary operator $\imult$ and
unit $\iunit$, an \emph{indexed comonad} over a category $\ccat$
comprises a family of object mappings $\iftor r$ where for all $r \in S$ and $A \in \obj{\ccat}$ then 
$\iftor r A \in \obj{\ccat}$ and:
%%
\begin{itemize}
\item a natural transformation $\counit_A : \iftor{\iunit} A \rightarrow A$, called the \emph{counit};
\item a family of mappings $\cobind{(-)}{r, s}$ from morphisms $\iftor
  r A \rightarrow B$ to \\ morphisms $\iftor{r \imult{} s} A
  \rightarrow \iftor s B$ in $\ccat$, natural in $A,B$, called
  \emph{coextend};
\end{itemize}
%
\noindent
such that for all $f : \iftor r \tau_1 \rightarrow \tau_2$ and $g : \iftor s \tau_2 \rightarrow \tau_3$ 
the following equations hold:
%
\renewcommand{\arraycolsep}{1.5em}
\begin{align*}
\begin{array}{ccc}
\counit \circ \cobind{f}{r, \iunit} = f & 
\cobind{{(\counit)}}{\iunit, r} = \ident{id} & 
\cobind{(g \circ \cobind{f}{r, s})}{(r \imult s), t} = \cobind{g}{s, t} \circ \cobind{f}{r, (s \imult t)} \\
\end{array}
\end{align*}
\end{definition}
%
The \emph{coextend} operation gives rise to an associative composition operation for
computations with contextual requirements (with \emph{counit} as the identity):
%
\[
\hat{\circ} : (\iftor r \tau_1 \rightarrow \tau_2) \rightarrow (\iftor s \tau_2 \rightarrow \tau_3) 
  \rightarrow (\iftor{r \imult s} \tau_1 \rightarrow \tau_3) \quad\quad\quad
g \, \hat{\circ} \, f = g \circ f^\dagger_{r,s}\quad
\]
%
The composition $\hat{\circ}$ best expresses the intention
of indexed comonads: contextual requirements of the composed
functions are combined. The properties of
the composition follow from the indexed comonad laws and the 
monoid $(S, \oplus, \iunit)$.

@@

For an equational theory of coeffects, consider first a notion 
of \emph{let}-binding equivalent to $(\lambda x . e_2)~e_1$, which
has the following type and coeffect rule:
%
\begin{equation}
\inference
  {\ctyp{s}{\Gamma} \vdash e_1 : \tau_1 &
   \ctyp{r_1 \wedge r_2}{(\Gamma, x : \tau_1)} \vdash e_2 : \tau_2}
  {\ctyp{r_1 \vee (r_2 \imult s)}{\Gamma} \vdash \kvd{let}~x = e_1~\kvd{in}~e_2
: \tau_2 }
\label{eq:let1}
\end{equation}

@@

The coeffect $r \vee (r \imult s)$ can
also be simplified for all our examples, leading to more intuitive
rules -- for implicit parameters $r \cup (r \cup s) = r \cup s$; for
liveness we get that $r \sqcup (r \sqcap s) = r$ and for dataflow we
obtain $\textit{max}(r, r+s) = r + s$.

>>>>

Successful programming language abstractions need to generalize a wide range of recurring
problems while capturing the key commonalities. These two aims are typically in opposition -- 
more general abstractions are less powerful, while less general abstractions cannot be
used as often.

In the previous chapter, we outlined a number of systems that capture how computations
access the environment in which they are executed. We identified two kinds of systems --
\emph{flat} capturing whole-context properties and \emph{structural} capturing per-variable
properties. As we show in Chapter~\ref{ch:unified}, the systems can be unified using a single 
abstraction. This is useful when implementing and composing the systems, but such abstraction 
is  \emph{less powerful} -- \ie~its generality hides useful properties that we can see 
when we consider the systems separately. For this reason, this and the next chapter discusses 
\emph{flat} and \emph{structural} systems separately.

@@

In the previous chapter, we looked at three important examples of systems that track whole-context 
properties. The type systems for whole-context liveness (Section~\ref{sec:applications-flat-live}) 
and whole-context data-flow (Section~\ref{sec:applications-flat-dataflow}) have a very similar 
structure -- their lambda abstraction duplicates the requirements and their application arises
from the combination of \emph{sequential} and \emph{point-wise} composition.

The system for tracking of implicit parameters (Section~\ref{sec:applications-flat-impl}), and
similar systems for rebindable resources, differ in two ways. In lambda abstraction, they split
the context requirements between the declaration-site and the call-site and they use only a single
operator on the indices, typically $\cup$.

\subsection{Contributions}

All of the examples are practically useful and important and so we want to be able to capture all
of them. Despite the differences, the systems can fit the same framework. The contributions of this
chapter are as follows:

\begin{itemize}
\item We present a \emph{flat coeffect calculus} with a type system that is parameterized by a 
  \emph{flat coeffect algebra} and can be instantiated to obtain all of the three examples
  discussed (Section~\ref{sec:flat-calculus}).

\item We give the equational theory of the calculus and discuss type-preservation for call-by-name
  and call-by-value reduction (Section~\ref{sec:flat-syntax}). We also extend the calculus
  with pairs and recursion (Section~\ref{sec:flat-exts}).

\item We present the semantics of the calculus in terms of \emph{indexed comonads}, which is a
  generalization of comonads, a category-theoretical dual of monads (Section~\ref{sec:flat-semantics}).
  The semantics provides deeper insight into how (and why) the calculus works.

\item We develop an alternative presentation of the system in terms of a simple structure 
  (semi-lattice) and use it to develop a type inference algorithm for flat coeffect calculus.
\end{itemize}

\subsection{Related work}

The development in this chapter can be seen as a counterpart to the well-known development of 
\emph{effect systems} \cite{effects-gifford} and the use of \emph{monads} \cite{monad-notions}
in programming languages. The syntax and type system of the flat coeffect calculus follows 
similar style as effect systems \cite{effects-polymorphic,effects-talpin-et-al}, but differs
in the structure, as explained in the previous chapter, most importantly in lambda abstraction.

Wadler and Thiemann famously show a correspondence between effect systems to monads 
\cite{monads-effects-marriage}, relating effectful functions $\tau_1 \xrightarrow{\sigma} \tau_2$ 
to monadic computations $M^\sigma \tau_1 \rightarrow \tau_2$. In this chapter, we show a similar
correspondence between \emph{coeffect systems} and \emph{comonads}. However, due to the asymmetry 
of $\lambda$-calculus, this is not a simple mechanical dualization.

The main purpose of the comonadic semantics presented in this chapter is to provide a semantic
motivation for the flat coeffect calculus. The semantics is inspired by the work of Uustalu and
Vene \cite{comonads-notions} who present the semantics of contextual computations (mainly for
data-flow) in terms of comonadic functions $C \tau_1 \rightarrow \tau_2$. Our \emph{indexed 
comonads} annotate the structure with information about the required context, \ie~$C^\sigma \tau_1 \rightarrow \tau_2$.
This is similar to the recent work on \emph{parameterized monads} by Katsumata \cite{monads-parametric}.

@@

The flat coeffect calculus is defined in terms of \emph{flat coeffect algebra}, which defines
the structure of context annotations, such as $\cclrd{r}, \cclrd{s}, \cclrd{t}$. These can be
sets of implicit parameters, integers or other values. The expressions of the calculus are those
of the $\lambda$-calculus with \emph{let} binding; assuming $T$ ranges over base types, the 
types of the calculus are defined as follows:
%
\begin{equation*}
\begin{array}{rcl}
e &::=& x \sep \lambda x.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
\tau &::=& T \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
\end{array}
\end{equation*}
%
We discuss pairs and recursion in Section~\ref{sec:flat-exts}. The type $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$
represents a function from $\tau_1$ to $\tau_2$ that requires additional context $\cclrd{r}$.
It can be viewed as a pure function that takes $\tau_1$ \emph{with} or \emph{wrapped in} a 
context $r$.

In the categorical semantics, the function $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$ is modelled
by a morphism $C^{\cclrd{r}} \tau_1 \rightarrow \tau_2$. However, the object $C^{\cclrd{r}}$
does not exist as a syntactical value. This is because we use comonads to define the 
\emph{semantics} rather than \emph{embedding} them into the language as in the meta-language
approaches (the distinction between the two approaches has been discussed in detail in 
Section~\ref{sec:path-sem-langs}). The annotations $\cclrd{r}$ are formed by an algebraic
structure discussed next.

@@

\subsection{Reconciling lambda abstraction}

Recall the lambda abstraction rules for the implicit parameters system (annotating the context
with sets of required parameters) and the data-flow system (annotating the context with the
number of past required values):
%
\begin{equation*}
\tyrule{abs-imp}
  {\coctx{\Gamma, x:\tau_1}{\aclrd{r} \cup \aclrd{s}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{s}} \tau_2 }
\;
\tyrule{abs-df}
  {\coctx{\Gamma, x:\tau_1}{\aclrd{n}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{n}} \tau_2 }
\end{equation*}

~

In order to capture both systems using a single calculus, we need a way of unifying the two
systems. For the data-flow system, this can be achieved by over-approximating the number of 
required past elements:
%
\begin{equation*}
\tyrule{abs-min}
  {\coctx{\Gamma, x:\tau_1}{\textnormal{min}(\aclrd{n}, \aclrd{m})} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{m}} \tau_2 }
\end{equation*}
%
The rule (\emph{abs-df}) is admissible in a system that includes the (\emph{abs-min}) rule. 
If we include sub-typing rule (on annotations of functions) and sub-coeffecting rule (on 
annotations of contexts), then the reverse is also true -- because 
$\textit{min}(\aclrd{n}, \aclrd{m}) \leq \aclrd{m}$ and $\textit{min}(\aclrd{n}, \aclrd{m}) \leq \aclrd{n}$.

\subsection{Flat coeffect algebra}
To make the flat coeffect system general enough, the algebra consists of three operations.
Two of them, $\cseq$ and $\cpar$, represent the \emph{sequential} and \emph{point-wise} composition, 
respectively and the third one, $\czip$ represents context \emph{merging}. The term merging should be 
understood semantically -- the operation models what happens when the semantics of lambda abstraction 
combines context available at the declaration-site and the call-site.

In addition to the three operations, we also require two special values used to annotate
variable access and constant access and a relation that defines the ordering.

@@

\noindent
In two of the three systems, some of the operators of the flat coeffect algebra coincide,
but the data-flow system requires all three. Similarly, the two special elements also 
coincide in some, but not all systems. The required laws are motivated by the aim to capture
common properties of the three examples, without unnecessarily restricting the system:

\begin{itemize}
\item The monoid $(\C, \cseq, \cunit)$ represents \emph{sequential} composition of (semantic)
functions. The laws of a monoid are required in order to form a category structure in the 
categorical model (Section~\ref{sec:flat-semantics}).

\item The monoid $(\C, \cpar, \czero)$ represents \emph{point-wise} composition, \ie~the case when the
same context is passed to multiple (independent) computations. The monoid laws guarantee 
that usual syntactic transformations on tuples and the unit value (Section~\ref{sec:flat-exts})
preserve the coeffect.

\item For the $\czip$ operation, we only require associativity in general. However, the operator can,
in some cases, form a monoid with the unit being the greatest element of the set. This 
alternative is discussed when we consider recursion (Section~\ref{sec:flat-exts}).
\end{itemize}

It is worth noting that the operators $\cpar$ and $\czip$ are dual in some of the systems. For 
example, in data-flow computations, they are \emph{max} and \emph{min} respectively. However, this
duality does not hold for implicit parameters. Using the syntactic reading, they represent 
\emph{merging} and \emph{splitting} of context requirements.

\paragraph{Ordering.}

The flat coeffect algebra requires a pre-order relation $\cleq$, which is used to define 
sub-coeffecting rule of the type system. When the monoid $(\C, \cpar, \czero)$ is idempotent
and commutative monoid (semi-lattice), the $\cleq$ relation can be defined in terms of $\cpar$ as:
%
\begin{equation*}
r \;\cleq\; s \;\Longleftrightarrow\; r \;\cpar\; s \;=\; s
\end{equation*}
%
This definition is consistent with all three examples that motivate flat coeffect calculus.
However, it cannot be used with the structural coeffects (where it fails for the bounded reuse 
calculus) and so we also do not use it for flat coeffects.

Furthermore, the $\cunit$ coeffect is often the top (greatest) or the bottom (smallest) 
element of the semi-lattice, but not in general. However, when this is the case, we are able
to prove certain properties of the calculus (Section~\ref{sec:syntax}).

\begin{figure}[t]
\begin{equation*}
\tyrule{var}
  {x : \tau \in \Gamma}
  {\coctx{\Gamma}{\cunit} \vdash x : \tau }
\end{equation*}
\begin{equation*}
\tyrule{const}
  {c : \tau \in \Delta}
  {\coctx{\Gamma}{\czero} \vdash c : \tau }
\end{equation*}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }
  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }\quad\quad(\cclrd{r'} \cleq \cclrd{r})
\end{equation*}
\begin{equation*}
\tyrule{app}
  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
   \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2}
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 &
    \coctx{\Gamma, x:\tau_1}{\cclrd{s}} \vdash e_2 : \tau_2}
  {\coctx{\Gamma}{\cclrd{s} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
\end{equation*}

\caption{Type system for the flat coeffect calculus}
\label{fig:flat-types}
\end{figure}

@@

\vspace{-0.4em}
\paragraph{Implicit parameters} use sets of names $S = \mathcal{P}(\ident{Id})$ as tags with
union $\cup$ for all three operators. Variable access is annotated with $\cunit = \emptyset$
and $\leq$ is subset ordering.

\vspace{-0.4em}
\paragraph{Liveness} uses a two point lattice $S=\{ \ident{D}, \ident{L} \}$ where
$\ident{D} \sqsubseteq \ident{L}$. Variables are annotated with the top element $e = \ident{L}$ 
and constants with bottom $\ident{D}$. The $\vee$ operation is $\sqcup$ (join) and $\wedge$
and $\cseq$ are both $\sqcap$ (meet).

\vspace{-0.4em}
\paragraph{Dataflow} tags are natural numbers $S = \mathbb{N}$ and operations
$\vee, \wedge$ and $\cseq$ correspond to \textit{max}, \textit{min} and $+$, respectively. 
Variable access is annotated with $\cunit = 0$ and the order $\leq$ is the standard ordering
of natural numbers.

@@

\section{Related work}

\subsection{What can be monad}
\label{sec:flat-related-monads}

@@

@@

% comonadic semantics
\newcommand{\iftor}[1]{C^{#1}}
\newcommand{\iftorhat}[1]{\hat{C}^{#1}}

@@

\newcommand{\cobind}[2]{#1^\dagger_{#2}}
\newcommand{\cmerge}[0]{ \ident{m} }
\newcommand{\csplit}[0]{ \ident{n} }
\newcommand{\counit}[0]{ \varepsilon }

@@

\vspace{-1em}
\subsection{Coeffect typing rules.}
Figure~\ref{fig:coeffect-typing} shows the
rules of the coeffect calculus, given some coeffect algebra $(S, \cseq, \vee, \wedge, \cunit)$.
The context required by a variable access (\emph{var}) is annotated with $\cunit$.
The sub-coeffecting rule (\emph{sub}) allows the contextual requirements of an expression 
to be generalized.

@@

In function application (\emph{app}), context requirements of both expressions and the 
function are combined as previously: the pointwise composition $\vee$ is used to combine
the coeffects of the expression representing a function $r$ and the coeffects of
the argument, sequentially composed with the coeffects of the function $s \cseq t$.

@@

\begin{figure}[t]
\vspace{-1em}
\begin{equation*}
\inference[(\emph{var})\;]
  {x : \tau \in \Gamma}
  {\ctyp{\cunit}{\Gamma} \vdash x : \tau }
%
\quad\quad
%
\inference[(\emph{app})\;]
  {\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{s}{\tau_1} \rightarrow \tau_2 &
   \ctyp{t}{\Gamma} \vdash e_2 : \tau_1 }
  {\ctyp{r \vee (s \cseq t)}{\Gamma} \vdash e_1~e_2 : \tau_2}
\end{equation*}
\begin{equation*}
\inference[(\emph{sub})\;]
  {\ctyp{s}{\Gamma} \vdash e : \tau }
  {\ctyp{r}{\Gamma} \vdash e : \tau }
\;(s\leq r)  
%
\quad\quad\quad\quad
%
\inference[(\emph{abs})\;]
  {\ctyp{r \wedge s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
  {\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
\end{equation*}
\caption{Type and coeffect system for the coeffect calculus}
\label{fig:coeffect-typing}
\vspace{-1em}
\end{figure}

@@

The indexed comonad structure provides a notion of composition
for computations with different contextual requirements.
%%
\begin{definition}
Given a monoid $(S, \cseq, \cunit)$ with binary operator $\cseq$ and
unit $\cunit$, an \emph{indexed comonad} over a category $\ccat$
comprises a family of object mappings $\iftor r$ where for all $r \in S$ and $A \in \obj{\ccat}$ then 
$\iftor r A \in \obj{\ccat}$ and:
%%
\begin{itemize}
\item a natural transformation $\counit_A : \iftor{\cunit} A \rightarrow A$, called the \emph{counit};
\item a family of mappings $\cobind{(-)}{r, s}$ from morphisms $\iftor
  r A \rightarrow B$ to \\ morphisms $\iftor{r \cseq{} s} A
  \rightarrow \iftor s B$ in $\ccat$, natural in $A,B$, called
  \emph{coextend};
\end{itemize}
%
\noindent
such that for all $f : \iftor r \tau_1 \rightarrow \tau_2$ and $g : \iftor s \tau_2 \rightarrow \tau_3$ 
the following equations hold:
%
\renewcommand{\arraycolsep}{1.5em}
\begin{align*}
\begin{array}{ccc}
\counit \circ \cobind{f}{r, \cunit} = f & 
\cobind{{(\counit)}}{\cunit, r} = \ident{id} & 
\cobind{(g \circ \cobind{f}{r, s})}{(r \cseq s), t} = \cobind{g}{s, t} \circ \cobind{f}{r, (s \cseq t)} \\
\end{array}
\end{align*}
\end{definition}
%
The \emph{coextend} operation gives rise to an associative composition operation for
computations with contextual requirements (with \emph{counit} as the identity):
%
\[
\hat{\circ} : (\iftor r \tau_1 \rightarrow \tau_2) \rightarrow (\iftor s \tau_2 \rightarrow \tau_3) 
  \rightarrow (\iftor{r \cseq s} \tau_1 \rightarrow \tau_3) \quad\quad\quad
g \, \hat{\circ} \, f = g \circ f^\dagger_{r,s}\quad
\]
%
The composition $\hat{\circ}$ best expresses the intention
of indexed comonads: contextual requirements of the composed
functions are combined. The properties of
the composition follow from the indexed comonad laws and the 
monoid $(S, \oplus, \cunit)$.

@@

For an equational theory of coeffects, consider first a notion 
of \emph{let}-binding equivalent to $(\lambda x . e_2)~e_1$, which
has the following type and coeffect rule:
%
\begin{equation}
\inference
  {\ctyp{s}{\Gamma} \vdash e_1 : \tau_1 &
   \ctyp{r_1 \wedge r_2}{(\Gamma, x : \tau_1)} \vdash e_2 : \tau_2}
  {\ctyp{r_1 \vee (r_2 \cseq s)}{\Gamma} \vdash \kvd{let}~x = e_1~\kvd{in}~e_2
: \tau_2 }
\label{eq:let1}
\end{equation}

@@

The coeffect $r \vee (r \cseq s)$ can
also be simplified for all our examples, leading to more intuitive
rules -- for implicit parameters $r \cup (r \cup s) = r \cup s$; for
liveness we get that $r \sqcup (r \sqcap s) = r$ and for dataflow we
obtain $\textit{max}(r, r+s) = r + s$.

Our calculus can be extended with \emph{let}-binding and \eqref{eq:let2}.
However, we also consider the cases when a
syntactic substitution $e_2[x \leftarrow e_1]$ has the coeffects
specified by the above rule \eqref{eq:let2} 
and prove \emph{subject reduction} theorem
for certain coeffect calculi.  We consider two common special cases
when the coeffect of variables $\cunit$ is the greatest ($\top$) or
least ($\bot$) element of the semi-lattice $(S, \vee)$ and derive
additional conditions that have to hold about the coeffect algebra:
